---
title: "eda_TCGA_Breast_cancer_ferroptosis"
author: "Ryan Rebernick"
date: "4/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description:
Analyzed TCGA breast cancer cohort using 306 gene ferroptosis gene signature. Overall see differences in ferroptosis levels by subtype. Differences in OS most pronounced in HER2-overexpressing breast cancer. 


# Load packages
```{r, warning=F, message=FALSE}

library('plyr')
library('dplyr')
library('stringr')
library('data.table')
library('tidyverse')
library('magrittr')

#BiocManager::install("singscore")
library('singscore')
#install.packages("ggpubr")
library('ggpubr')
library('ggfortify')
library('cowplot')
library(survival)
library(matrixStats)
library(biomaRt)
library(EPIC)

```



# Load data
Incorporates metadata from 2 UCSC Xena entries in order to capture information on ER/PR/Her2 receptor status published in PMID: 23000897. 
The primary UCSC dataset i'm using: GDC_TCGA_Breast_cancer_BRCA (N=1217)
The UCSC dataset with information on HR status: TCGA_Breast_cancer_BRCA (n=1218)
```{r}

# fpkm
fpkm <- fread('/mctp/share/users/rebernrj/projects/ferroptosisGeneSig/data/ucsc_xena/datasets/GDC_TCGA_Breast_cancer_BRCA/TCGA-BRCA.htseq_fpkm.tsv', data.table = F)

# phenotypic data
pheno <- fread('/mctp/share/users/rebernrj/projects/ferroptosisGeneSig/data/ucsc_xena/datasets/GDC_TCGA_Breast_cancer_BRCA/TCGA-BRCA.GDC_phenotype.tsv', data.table = F)

# survival data
surv <- fread('/mctp/share/users/rebernrj/projects/ferroptosisGeneSig/data/ucsc_xena/datasets/GDC_TCGA_Breast_cancer_BRCA/TCGA-BRCA.survival.tsv', data.table = F)

# load updated metadata information from other UCSC xena dataset that uses TCGA data
# The reason this is used is b/c has info on ER/PR/Her2 status from PMID: 23000897. The associated expression data from that dataset is not used b/c has a normalized expression matrix rather than fpkm
otherMeta <- fread('/mctp/share/users/rebernrj/projects/ferroptosisGeneSig/data/ucsc_xena/datasets/TCGA_Breast_cancer_BRCA/TCGA.BRCA.sampleMap%2FBRCA_clinicalMatrix', data.table = F)


```



# Prepare data for gene signature analysis

### Get Common names from biomart for use in signature
```{r}

# Get biomart information
ensembl <- useMart("ensembl")
ensembl = useDataset("hsapiens_gene_ensembl", mart=ensembl)


key = getBM(attributes=c('ensembl_gene_id', 'external_gene_name'), 
      filters = 'chromosome_name', 
      values = c(1:22, 'X', 'Y'), 
      mart = ensembl,
      useCache = FALSE)

```

### Convert database ensembl ids to common names
```{r}

# Trim off version information of ensembl names
fpkm$Ensembl_ID <- gsub('\\.[0-9]*', '', fpkm$Ensembl_ID)

# merge with common names
fpkm <- merge(key, fpkm, by.x = 'ensembl_gene_id', by.y='Ensembl_ID')

# determine if duplicates are in gene signature
sig306 <- fread('/mctp/share/users/rebernrj/projects/ferroptosisGeneSig/out/gene_signatures/erastin_clue24h_gse104462_306genes.txt',data.table = F)
table(fpkm$external_gene_name[duplicated(fpkm$external_gene_name)] %in% sig306$V1)

# remove duplicate genes since not in gene signature
fpkm <- fpkm[!duplicated(fpkm$external_gene_name),]

# set genes as rownames
rownames(fpkm) <- fpkm$external_gene_name
fpkm %<>% dplyr::select(-ensembl_gene_id, -external_gene_name)

```

### eliminate genes which are <1 fpkm in over 50% of samples
```{r}

# get 50th percentile value
fpkm$quart2 <- apply(fpkm, MARGIN = 1, quantile, na.rm = T, probs=0.5)

# filter
fpkm %<>% dplyr::filter(quart2 > 0) %>%
  dplyr::select(-quart2)

```

### Combine old metadata with new metadata
This is where the otherMetadata from a different UCSC Xena TCGA dataset comes in.
Samples with end of 11 are normal tissue
```{r}

# Create generic sample ID column for merging with other metadata
pheno %<>% mutate(genericID = substr(submitter_id.samples, 1, nchar(submitter_id.samples)-1))

# merge; keep all x samples otherwise will lose the normal tissue
pheno <- merge(pheno, otherMeta, by.x = 'genericID', by.y = 'sampleID', all.x = T)

# filter uneeded cols from other metadata
pheno %<>% dplyr::select(!contains('.y'))

```


### Merge phenotypic/survival data with fpkm data
```{r}

# merge phenotypic and fpkm data
temp <- fpkm %>% t() %>% as.data.frame() %>% cbind(comp = 'ALL', .)

brca <- merge(pheno, surv, by.x = 'submitter_id.samples', by.y='sample')
brca <- merge(brca, temp, by.x = 'submitter_id.samples', by.y=0)


```



# EPIC (cell proportion estimate)

### Run EPIC and combine with existing metadata
Triggers error regarding convergance. Documentation indicates this might happen. 
```{r}

# Run epic
epic <- EPIC(bulk = fpkm,
            withOtherCells = T)


# Merge dataframes for downstream use
brca <- merge(brca, epic$cellFractions, by.x = 'submitter_id.samples', by.y = 0)

# Order dataframe
brca %<>% set_rownames(brca$Row.names) %>%
  dplyr::relocate(Bcells:otherCells, .after=vital_status)

```



# Select unique samples 
There are duplicates with some patients having multiple tissue types (sample_type.samples).
Some patients have duplicate values of the same tissue types (patient = 2672). The difference between these samples is the variable is_ffpe.samples, whether the sample was taken ffpe or flash frozen. I stratified for taking samples that are flash frozen when possible.
```{r}

table(brca$sample_type.samples)

# arrange df to allow for easy duplicate removal
brcauniq <- brca %>% 
  ## set sample type to be a factor to allow for sorting
  mutate(sample_type.samples = factor(brca$sample_type.samples, levels = c('Primary Tumor', 'Metastatic', 'Solid Tissue Normal'))) %>%
  ## arrange based on sample type and whether ffpe
  arrange(sample_type.samples, is_ffpe.samples)

# remove the duplicates based on sorting above
brcauniq <- brcauniq[!duplicated(brcauniq$patient_id.x),]

# verify patients and ffpe are uniform
table(brcauniq$is_ffpe.samples)
table(duplicated(brcauniq$patient_id.x))

```



# General QC of brca cohort

### PCA by sample origin
```{r}

# All genes and all samples
set.seed(11)
brca$comp <- brca$sample_type.samples
dat <- brca %>% dplyr::select(comp, TSPAN6:ncol(brca))
autoplot(dat %>% dplyr::select(-comp) %>% prcomp(),
              data = dat,
              colour = 'comp',
              main = paste0('PCA of all genes split by sample type')) +
  theme_cowplot(12)


```



# Create functions neeed to analyze gene signatures

### Create a function to return a signature scores using 3 methods:
1) Sample scoring package - less dependence of score on other samples
https://bioconductor.org/packages/devel/bioc/vignettes/singscore/inst/doc/singscore.html#21_Sample_scoring
2) Nondirectional Summation of Z scores
- Sum the absolute values of Z scores across up and downregulated genes
3) Directional Summation of Z scores
- Sum z scores of upregulated genes + -1*(Sum of Z scores of downregulated genes)
```{r}

# DF - the cleaned dataframe from above
# UPGENES - list of genes upregulated in gene signature
# DOWNGENES - list of genes downregulated in gene signature
# METACOLS - the columns of metadata within DF (needs to be one greater than number of metacols currently; d/t Row.names being added during merge)

calculate_signature_scores <- function(DF, UPGENES, DOWNGENES, METACOLS = 1:292){
  
  # create temp dataframe
  cnts <- DF %>% set_rownames(paste0('Sample_', 1:length(DF[,1])))
  
  # Score samples using singScore and merge
  ## score
  rankData <- cnts %>% dplyr::select(-METACOLS) %>%  t() %>% rankGenes()
  if(!is.null(UPGENES) & !is.null(DOWNGENES)){
    scoredf <- simpleScore(rankData, upSet = UPGENES, downSet = DOWNGENES)
  } else if(is.null(UPGENES) & !is.null(DOWNGENES)) {
    scoredf <- simpleScore(rankData, downSet = DOWNGENES)
    print('Only DOWN geneset is valid')
  } else if(!is.null(UPGENES) & is.null(DOWNGENES)) {
    scoredf <- simpleScore(rankData, upSet = UPGENES)
    print('Only UP geneset is valid')
  } else {
    print('Both gene sets NULL')
    break
  }
  nms <- paste0('ss.', colnames(scoredf))
  colnames(scoredf) <- nms
  ## merge SingScore Total score with tmp df
  cnts <- merge(cnts, scoredf, by = 0)
  
  
  # Calcluate nondirectional z scores 
  temp <- apply(cnts[,c(UPGENES, DOWNGENES)], 2, scale)
  temp$nondirectional_Zsum <- rowSums(abs(temp))
  ## Add Z score to dataframe 
  cnts$nondirectional_Zsum <- temp$nondirectional_Zsum
  
  
  # Calculate directional Z scores
  ## Upregulated genes
  temp <- apply(cnts %>% dplyr::select(all_of(UPGENES)), 2, scale)
  if(!is_empty(temp)) {
    temp$upZsum <- rowSums(temp)
    ## Add Z score to dataframe 
    cnts$upZsum <- temp$upZsum
  } else {
    cnts$upZsum <- 0
  }
  ## Downregulated genes
  temp <- apply(cnts %>% dplyr::select(all_of(DOWNGENES)), 2, scale)
  if(!is_empty(temp)) {
    temp$downZsum <- rowSums(temp)
    ## Add Z score to dataframe 
    cnts$downZsum <- temp$downZsum
  } else {
    cnts$downZsum <- 0
  }
  
  # Select columns
  cnts %<>% dplyr::select(METACOLS, all_of(nms), upZsum, downZsum)
  cnts %<>% mutate(directional_Zsum = upZsum + (downZsum * -1)) %>%
    dplyr::select(-Row.names)
  
  # Correlation
  print(paste0('Correlation b/t SingScore and Directional Zsum: ', cor(cnts$ss.TotalScore, cnts$directional_Zsum)))
  
  # Return
  return(cnts)
  
  
}

```

### Function to plot comparisons of interest
```{r}

# DF - dataframe to use; nbl format
# UPGENES - list of upregulated genes
# DOWNGENES - list of downregulated genes
# COMP - comparison variable/column
# STAT_TEST - c('wilcox.test', 't.test', 'anova', 'kruskal.test')
# TIMEPOINT = timepoint in nbl cohort to look at
make_plots <- function(
  DF,
  UPGENES, 
  DOWNGENES, 
  COMP='comp', 
  STAT_TEST = 'wilcox.test',
  TITLE = 'XX gene sig'){
  
  ####################
  # return genes not in provided gene list and remove from query
  ####################
  print(paste(c('Up genes not in count matrix:', UPGENES[!UPGENES %in% colnames(DF)]), collapse = ' '))
  print(paste(c('Down genes not in count matrix:', DOWNGENES[!DOWNGENES %in% colnames(DF)]), collapse = ' '))
  UPGENES <- UPGENES[UPGENES %in% colnames(DF)]
  DOWNGENES <- DOWNGENES[DOWNGENES %in% colnames(DF)]
  
  ####################
  # PCA PLOT by group
  ####################
  set.seed(11)
  dat <- DF[, c(COMP, UPGENES, DOWNGENES)]
  a <- autoplot(dat %>% dplyr::select(-COMP) %>% prcomp(),
           data = dat,
           colour = COMP,
           main = paste0('PCA using ', TITLE)) +
    theme_cowplot(12)
  
  
  ####################
  # Calclulate scores
  ####################
  scores <- calculate_signature_scores(DF = DF, 
                                   UPGENES = UPGENES,
                                   DOWNGENES = DOWNGENES)
  
  ####################
  # Boxplot 
  ####################

  # # select samples
  # if(TIMEPOINT %in% c('dx2prim', 'dxmet', 'dxprim', 'posttx', 'rel_ref')){
  #   DF %<>% dplyr::filter(timepoint %in% c(TIMEPOINT))
  # } else if(TIMEPOINT == 'all'){
  #   print('All timepoints included')
  # } else {
  #   break
  # }
  
  # Plot 
  b1 <- scores %>% ggplot(aes(x=get(COMP), y=ss.TotalScore)) + 
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(alpha=0.6, width=0.15) +
    theme_bw(base_size=16) +
    labs(subtitle=TITLE) +
    stat_compare_means(method = STAT_TEST) +
    xlab(NULL)
  
  # Plot 
  # b2 <- scores %>% ggplot(aes(x=get(COMP), y=nondirectional_Zsum)) + 
  #   geom_boxplot() +
  #   geom_jitter(alpha=0.6, width=0.15) +
  #   theme_bw(base_size=16) +
  #   labs(subtitle="Nondirectional Z score by Risk group in dxprim") +
  #   stat_compare_means(method = STAT_TEST) +
  #   xlab(NULL)
  
  # Plot 
  b3 <- scores %>% ggplot(aes(x=get(COMP), y=directional_Zsum)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(alpha=0.6, width=0.15) +
    theme_bw(base_size=16) +
    labs(subtitle=TITLE) +
    stat_compare_means(method = STAT_TEST) +
    xlab(NULL) +
    theme(axis.text.x=element_text(angle = -90, hjust = 0))
  
  return(list(scores,a,b1,b3))
  
}

```



# Analyze gene signatures 

## 306 gene signature

### Load gene signature
```{r}
sig306 <- fread('/mctp/share/users/rebernrj/projects/ferroptosisGeneSig/out/gene_signatures/erastin_clue24h_gse104462_306genes.txt',
                data.table = F)

# Upregulated
sig306_up <- sig306 %>% 
  dplyr::filter(upregulated == T) %>% 
  dplyr::select(V1) %>% unlist()

# Downregulated
sig306_dn <- sig306 %>% 
  dplyr::filter(upregulated == F) %>% 
  dplyr::select(V1) %>% unlist()


```

### Distribution of 306 gene signature scores in all patients. Correlation with SLC levels
Also check correlation with MIR522 -secreted by fibroblasts and inhibits ferroptosis (https://molecular-cancer.biomedcentral.com/articles/10.1186/s12943-020-01168-8#:~:text=Ferroptosis%20is%20a%20novel%20mode,various%20bioactive%20substances%2C%20including%20exosomes)
Check correlation of other genes associated with ferroptosis too
```{r}

# Define comparison
brcauniq$comp <- 'all'

# get PCA and boxplots
cur <- make_plots(
  DF = brcauniq,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'anova',
  TITLE = '306 gene sig')

# Plot distribution of sing score
cur[3]
#cur[4]

# Look at correlation of singscore with SLC levels
cur <- cur[[1]]
cur %<>% dplyr::select(submitter_id.samples, ss.TotalScore, directional_Zsum) %>%
  merge(., brcauniq, by = 'submitter_id.samples')
## SLC3A2 expression
ggscatter(cur, x='ss.TotalScore', y='SLC3A2',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")
## SLC7A11
ggscatter(cur, x='ss.TotalScore', y='SLC7A11',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")

# SLC7A11 vs SLC3A2
ggscatter(cur, x='SLC3A2', y='SLC7A11',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")

# Other genes of interest with gs expression
## ALOX15 expression
ggscatter(cur, x='ss.TotalScore', y='ALOX15',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")
## ALOX15B expression
ggscatter(cur, x='ss.TotalScore', y='ALOX15B',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")
## PLIN2 expression
ggscatter(cur, x='ss.TotalScore', y='PLIN2',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")
## GDF15 expression
ggscatter(cur, x='ss.TotalScore', y='GDF15',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")

# Immune cell level with gene level
## CAFs with Alox15
ggscatter(cur, x='CAFs', y='ALOX15B',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")


```

### Correlation with cell proportions
```{r}

# Load gene sig scores
cur <- make_plots(
  DF = brca,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'anova',
  TITLE = '306 gene sig')
cur <- cur[[1]]


# CD4
ggscatter(cur, x='ss.TotalScore', y='CD4_Tcells',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")

# CD8
ggscatter(cur, x='ss.TotalScore', y='CD8_Tcells',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")

# B cells
ggscatter(cur, x='ss.TotalScore', y='Bcells',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")

# CAFs
ggscatter(cur, x='ss.TotalScore', y='CAFs',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")

# Endothelial
ggscatter(cur, x='ss.TotalScore', y='Endothelial',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")

# Macrophages
ggscatter(cur, x='ss.TotalScore', y='Macrophages',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")

# NKcells
ggscatter(cur, x='ss.TotalScore', y='NKcells',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")

# otherCells
ggscatter(cur, x='ss.TotalScore', y='otherCells',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")

```

### Correlation of 306gs with elife score
```{r}

# Load signature
elife_up <- c('CHAC1', 'DDIT4', 'LOC284561', 'ASNS', 'TSC22D3', 'DDIT3', 'JDP2', 'SESN2',
              'SLC1A4', 'PCK2', 'SLC7A11', 'TXNIP', 'VLDLR', 'GPT2', 'PSAT1', 'C9ORF150',
              'SLC7A5', 'HERPUD1', 'XBP1', 'ATF3', 'SLC3A2', 'CBS', 'ATF4', 'ZNF419', 
              'KLHL24', 'TRIB3', 'ZNF643', 'ATP6V1G2', 'VEGFA', 'GDF15', 'TUBE1', 
              'ARRDC3', 'CEBPG')
elife_down <- c('SNORA16A', 'RGS4', 'MUTED-TXNDC5', 'LOC390705')


# Determine correlation
## get elife values
brca$comp <- 'all'
cur <- make_plots(
  DF = brca,
  UPGENES = elife_up, 
  DOWNGENES = elife_down, 
  COMP = 'comp', 
  STAT_TEST = 'anova',
  TITLE = 'elife gene sig')
cur <- cur[[1]]
cur %<>% dplyr::select(submitter_id.samples, ss.TotalScore, directional_Zsum) %>% set_colnames(c('submitter_id.samples', 'elife.ss', 'elife.zsum'))
## get sig306 values
temp <- make_plots(
  DF = brca,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'anova',
  TITLE = 'elife gene sig')
temp <- temp[[1]]
temp %<>% dplyr::select(submitter_id.samples, ss.TotalScore, directional_Zsum) %>% set_colnames(c('submitter_id.samples', 'sig306.ss', 'sig306.zsum'))
##merge
cur %<>% merge(., temp, by = 'submitter_id.samples')
## Plot singscore elife vs singscore 306
ggscatter(cur, x='elife.ss', y='sig306.ss',
          add='reg.line',
          add.params = list(color='blue', fill='lightgray'),
          conf.int = T) +
  stat_cor(method = "pearson")
## Determine if the high scoring samples are the same in both scores
cur %<>% 
  mutate(elife_sub = ifelse(elife.ss >= quantile(elife.ss)[4], 'elife_hi', 'elife_lo')) %>%
  mutate(sig306_sub = ifelse(sig306.ss >= quantile(sig306.ss)[4], 'sig306_hi', 'sig306_lo'))
table(cur$elife_sub, cur$sig306_sub)


```

### 306gs with Breast cancer mutational status
```{r}

# Define comparison
dat <- brcauniq %>% dplyr::mutate(comp = PAM50_mRNA_nature2012)
dat %<>% dplyr::filter(comp %in% c('Basal-like', 'HER2-enriched', 'Luminal A', 'Luminal B', 'Normal-like')) %>%
  mutate(comp = factor(comp, levels = c('Normal-like', 'Luminal A', 'Luminal B', 'HER2-enriched', 'Basal-like')))

# plot PCA and boxplots
cur <- make_plots(
  DF = dat,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'anova',
  TITLE = '306gs: Breast cancer PAM50 classification')

# plot and get df

cur[3]
#cur[4]
```

### 306gs with Breast cancer Her2 status
```{r}

# Define comparison
dat <- brcauniq %>% dplyr::mutate(comp = HER2_Final_Status_nature2012)
dat %<>% dplyr::filter(comp %in% c('Negative', 'Positive'))

# plot PCA and boxplots
cur <- make_plots(
  DF = dat,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'wilcox.test',
  TITLE = '306gs: Breast cancer Her2 classification')

# plot and get df

cur[3]
#cur[4]
```

### 306gs with Breast cancer PR status
```{r}

# Define comparison
dat <- brcauniq %>% dplyr::mutate(comp = PR_Status_nature2012)
dat %<>% dplyr::filter(comp %in% c('Negative', 'Positive'))

# plot PCA and boxplots
cur <- make_plots(
  DF = dat,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'wilcox.test',
  TITLE = '306gs: Breast cancer PR classification')

# plot and get df

cur[3]
#cur[4]
```

### 306gs with Breast cancer ER status
```{r}

# Define comparison
dat <- brcauniq %>% dplyr::mutate(comp = ER_Status_nature2012)
dat %<>% dplyr::filter(comp %in% c('Negative', 'Positive'))

# plot PCA and boxplots
cur <- make_plots(
  DF = dat,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'wilcox.test',
  TITLE = '306gs: Breast cancer ER classification')

# plot and get df

cur[3]
#cur[4]
```

### 306 gene signature for tissue type
```{r}

# Define comparison
brca$comp <- brca$sample_type.samples
dat <- brca %>% dplyr::filter(sample_type.samples %in% c('Metastatic', 'Primary Tumor', 'Solid Tissue Normal'))

# plot PCA and boxplots
cur <- make_plots(
  DF = dat,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'anova',
  TITLE = '306gs: Tissue type')

# plot and get df

cur[3]
#cur[4]

```

### Examine variables by table fxn to help narrow down which are interesting
discrete: Metastasis_nature2012 radiation_therapy.x menopause_status.x
continuous: 
```{r}

# get sigscores 
cur <- make_plots(
  DF = brcauniq,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'anova',
  TITLE = 'NA')
cur <- cur[[1]]

# stratify by sigscores quartile 
cur %<>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[3], 'gs_quant_3', 'gs_quant_2')) %>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[4], 'gs_quant_4', comp)) %>%
  mutate(comp = ifelse(ss.TotalScore < quantile(ss.TotalScore)[2], 'gs_quant_1', comp)) 


# Discrete variables 
disc <- c('Tumor_nature2012', 'Metastasis_nature2012', 'radiation_therapy.x', 'menopause_status.x')

# loop through discrete variables
for(x in disc){
  print(x)
  print(table(cur$comp, cur[,x]))
}


# Continuous variables
cont <- c(NULL)

# loop through continuous variables
for(x in cont){
  
  print(x)

  # create quartiles of current continuous variable
  cur$var <- 'quart3'
  cur$var <- ifelse(cur[,x] >= quantile(cur[,x], na.rm = T)[4], 'quart4', cur$var)
  cur$var <- ifelse(cur[,x] <= quantile(cur[,x], na.rm = T)[3], 'quart2', cur$var)
  cur$var <- ifelse(cur[,x] <= quantile(cur[,x], na.rm = T)[2], 'quart1', cur$var)

  # table
  print(table(cur$comp, cur$var))
  
}


```

### 306 gene signature for pathologic_stage
```{r}

# Define comparison
dat <- brcauniq %>% 
  mutate(comp = ifelse(pathologic_stage %in% c('Stage I', 'Stage IA', 'Stage IB'), 'Stage 1', pathologic_stage)) %>% 
  mutate(comp = ifelse(pathologic_stage %in% c('Stage II', 'Stage IIA', 'Stage IIB', 'Stage IIC'), 'Stage 2', comp)) %>%
  mutate(comp = ifelse(pathologic_stage %in% c('Stage III', 'Stage IIIA', 'Stage IIIB', 'Stage IIIC'), 'Stage 3', comp)) %>%
  mutate(comp = ifelse(pathologic_stage %in% c('Stage IV', 'Stage IVA', 'Stage IVb'), 'Stage 4', comp)) %>%
  dplyr::filter(!comp %in% c('[Discrepancy]', '', 'Stage X'))
                                

# plot PCA and boxplots
cur <- make_plots(
  DF = dat,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'anova',
  TITLE = '306gs: tumor stage')

# plot
cur[3]


```


### 306 gene signature for Metastasis_nature2012
```{r}

# Define comparison
brca$comp <- brca$Metastasis_nature2012
dat <- brca %>% dplyr::filter(comp %in% c('M0', 'M1'))

# plot PCA and boxplots
cur <- make_plots(
  DF = dat,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'wilcox.test',
  TITLE = '306gs: Metastasis_nature2012')

# plot and get df

cur[3]
#cur[4]

```

### 306 gene signature for radiation_therapy.x
```{r}

# Define comparison
brca$comp <- brca$radiation_therapy.x
dat <- brca %>% dplyr::filter(comp %in% c('NO', 'YES'))

# plot PCA and boxplots
cur <- make_plots(
  DF = dat,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'wilcox.test',
  TITLE = '306gs: radiation_therapy.x')

# plot and get df

cur[3]
#cur[4]

```

### 306 gene signature for menopause_status.x
```{r}

# Define comparison
brca$comp <- brca$menopause_status.x
dat <- brca %>% dplyr::filter(comp %in% c('Post (prior bilateral ovariectomy OR >12 mo since LMP with no prior hysterectomy)', 'Pre (<6 months since LMP AND no prior bilateral ovariectomy AND not on estrogen replacement)'))
dat %<>% mutate(comp = ifelse(comp == 'Pre (<6 months since LMP AND no prior bilateral ovariectomy AND not on estrogen replacement)', 'pre', 'post'))

# plot PCA and boxplots
cur <- make_plots(
  DF = dat,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'wilcox.test',
  TITLE = '306gs: menopause_status.x')

# plot and get df

cur[3]
#cur[4]

```

### 306 gene signature Kaplan meijer for OS - 1st vs 4th quartile
Her2+ worse than luminal B worse than Luminal A == Basal
```{r}

# temp df to hold filtered skcmuniq data
dat <- brcauniq 

# get sigscores for stratifying in kaplan meijer curve
cur <- make_plots(
  DF = dat,
  UPGENES = sig306_up, 
  DOWNGENES = sig306_dn, 
  COMP = 'comp', 
  STAT_TEST = 'anova',
  TITLE = '306 gene sig')
cur <- cur[[1]]

# plot Kaplan meijer by subtype
temp <- cur %>% dplyr::filter(!PAM50_mRNA_nature2012 %in% c('Normal-like', ''))
temp %<>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[3], 3, 2)) %>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[4], 4, comp)) %>%
  mutate(comp = ifelse(ss.TotalScore < quantile(ss.TotalScore)[2], 1, comp)) 
temp %<>% dplyr::filter(comp %in% c(1,4))
km_fit <- survfit(Surv(OS.time, rep(1, length(temp[,1]))) ~ PAM50_mRNA_nature2012, data=temp)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit,
         main = 'All subtypes stratified by subtype',
         conf.int = F)

# plot Kaplan meijer for all subtypes by 306gs median
temp <- cur 
temp %<>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[3], '3/4', '1/2'))
km_fit <- survfit(Surv(OS.time, rep(1, length(temp[,1]))) ~ comp, data=temp)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit,
         main = 'All subtypes stratified by 306gs score mean',
         conf.int = F)

# plot Kaplan meijer for all subtypes by 306gs visual cutoff
temp <- cur 
temp %<>%
  mutate(comp = ifelse(ss.TotalScore >= -0.08, 'High-ferrop', 'Low-ferrop'))
km_fit <- survfit(Surv(OS.time, rep(1, length(temp[,1]))) ~ comp, data=temp)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit,
         main = 'All subtypes stratified by 306gs visual cutoff (-0.08)',
         conf.int = F)

# plot Kaplan meijer for all subtypes by 306gs
temp <- cur 
temp %<>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[3], 3, 2)) %>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[4], 4, comp)) %>%
  mutate(comp = ifelse(ss.TotalScore < quantile(ss.TotalScore)[2], 1, comp)) 
#temp %<>% dplyr::filter(comp %in% c(1,4))
km_fit <- survfit(Surv(OS.time, rep(1, length(temp[,1]))) ~ comp, data=temp)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit,
         main = 'All subtypes stratified by 1st/4th 306gs score quartile',
         conf.int = F)


# plot Kaplan meijer for basal
temp <- cur %>% dplyr::filter(PAM50_mRNA_nature2012 %in% c('Basal-like'))
temp %<>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[3], 3, 2)) %>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[4], 4, comp)) %>%
  mutate(comp = ifelse(ss.TotalScore < quantile(ss.TotalScore)[2], 1, comp)) 
#temp %<>% dplyr::filter(comp %in% c(1,4))
km_fit <- survfit(Surv(OS.time, rep(1, length(temp[,1]))) ~ comp, data=temp)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(km_fit,
         main = 'Basal-like stratified by 306gs score quartile',
         conf.int = F)

# plot Kaplan meijer for HER2-enriched
temp <- cur %>% dplyr::filter(PAM50_mRNA_nature2012 %in% c('HER2-enriched'))
temp %<>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[3], 3, 2)) %>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[4], 4, comp)) %>%
  mutate(comp = ifelse(ss.TotalScore < quantile(ss.TotalScore)[2], 1, comp)) 
#temp %<>% dplyr::filter(comp %in% c(1,4))
km_fit <- survfit(Surv(OS.time, rep(1, length(temp[,1]))) ~ comp, data=temp)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(object = km_fit, 
         main = 'HER2-enriched stratified by 306gs score quartile',
         conf.int = F)

# plot Kaplan meijer for Luminal A
temp <- cur %>% dplyr::filter(PAM50_mRNA_nature2012 %in% c('Luminal A'))
temp %<>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[3], 3, 2)) %>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[4], 4, comp)) %>%
  mutate(comp = ifelse(ss.TotalScore < quantile(ss.TotalScore)[2], 1, comp)) 
#temp %<>% dplyr::filter(comp %in% c(1,4))
km_fit <- survfit(Surv(OS.time, rep(1, length(temp[,1]))) ~ comp, data=temp)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(object = km_fit, 
         main = 'Luminal A stratified by 306gs score quartile',
         conf.int = F)

# plot Kaplan meijer for Luminal B
temp <- cur %>% dplyr::filter(PAM50_mRNA_nature2012 %in% c('Luminal B'))
temp %<>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[3], 3, 2)) %>%
  mutate(comp = ifelse(ss.TotalScore >= quantile(ss.TotalScore)[4], 4, comp)) %>%
  mutate(comp = ifelse(ss.TotalScore < quantile(ss.TotalScore)[2], 1, comp)) 
#temp %<>% dplyr::filter(comp %in% c(1,4))
km_fit <- survfit(Surv(OS.time, rep(1, length(temp[,1]))) ~ comp, data=temp)
summary(km_fit, times = c(1,30,60,90*(1:10)))
autoplot(object = km_fit, 
         main = 'Luminal B stratified by 306gs score quartile',
          conf.int = F)
    
```
