---
title: "eda_nbl_cohort"
author: "Ryan Rebernick"
date: "4/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Description:


# Load packages
```{r, warning=F, message=FALSE}

library('plyr')
library('dplyr')
library('stringr')
library('data.table')
library('tidyverse')
library('magrittr')

#BiocManager::install("singscore")
library('singscore')

```


# Load NBL cohort data and select samples for analysis
## Load NBL counts data and merge with metadata
17 duplicate gene names are present. These are just removed. If they are needed for any of the gene signatures they can be added to the dataset later. 
```{r}

# Load RNA-seq counts
nbl <- fread('/mctp/share/users/rebernrj/projects/ferroptosisGeneSig/data/nbl_cohort/release/ForRyan/nbl_expression-cts.csv', data.table = F)

# Remove any duplicate gene_ids so can transpose matrix
## View duplicate common gene_ids
duplicates <- nbl$gene_name[nbl$gene_name %>% duplicated()]
## remove duplicates
nbl %<>% dplyr::filter(!gene_name %in% duplicates)

# reformat the RNA-seq counts for merging with the clinical data
nbl %<>% set_rownames(nbl$gene_name) %>%
  dplyr::select(-gene_name, -gene_id) %>%
  t() %>%
  as.data.frame()

# Load the clinical data
temp <- fread('/mctp/share/users/rebernrj/projects/ferroptosisGeneSig/data/nbl_cohort/release/ForRyan/20210403_ForRyan_2021_0403_nbl_rnaseq.txt', data.table = F)

# Merge dataframes for downstream use
nbl <- merge(temp, nbl, by.x = 'rlib', by.y = 0)


```

## Select samples needed
```{r}


nbl %<>% dplyr::filter(r.capture == 'capt')







```


# Analyze gene signatures
## Create a function to return a signature scores using 3 methods:
1) Sample scoring package - less dependence of score on other samples
https://bioconductor.org/packages/devel/bioc/vignettes/singscore/inst/doc/singscore.html#21_Sample_scoring
2) Nondirectional Summation of Z scores
- Sum the absolute values of Z scores across up and downregulated genes
3) Directional Summation of Z scores
- Sum z scores of upregulated genes + -1*(Sum of Z scores of downregulated genes)
```{r}

# nbl_df - the cleaned nbl dataframe from above
# upGenes - list of genes upregulated in gene signature
# downGenes - list of genes downregulated in gene signature
# metacols - the columns of metadata within nbl_df

calculate_signature_scores <- function(nbl_df, upGenes, downGenes, metacols = 1:28){
  
  # create temp dataframe
  cnts <- nbl_df %>% set_rownames(paste0('Sample_', 1:length(cnts$rlib)))
  
  # return genes not in provided gene list and remove from query
  print(paste0('Up genes not in count matrix: ', upGenes[!upGenes %in% colnames(cnts)]))
  print(paste0('Down genes not in count matrix: ', downGenes[!downGenes %in% colnames(cnts)]))
  upGenes <- upGenes[upGenes %in% colnames(cnts)]
  downGenes <- downGenes[downGenes %in% colnames(cnts)]
  
  
  # Score samples using singScore and merge
  ## score
  rankData <- cnts %>% dplyr::select(-metacols) %>%  t() %>% rankGenes()
  scoredf <- simpleScore(rankData, upSet = upGenes, downSet = downGenes)
  colnames(scoredf) <- paste0('ss.', colnames(scoredf))
  ## merge SingScore Total score with tmp df
  cnts <- merge(cnts, scoredf, by = 0)
  
  
  # Calcluate nondirectional z scores 
  temp <- apply(cnts %>% dplyr::select(all_of(upGenes), all_of(downGenes)), 2, scale)
  temp$nondirectional_Zsum <- rowSums(abs(temp))
  ## Add Z score to dataframe 
  cnts$nondirectional_Zsum <- temp$nondirectional_Zsum
  
  
  # Calculate directional Z scores
  ## Upregulated genes
  temp <- apply(cnts %>% dplyr::select(all_of(upGenes)), 2, scale)
  temp$upZsum <- rowSums(temp)
  ## Add Z score to dataframe 
  cnts$upZsum <- temp$upZsum
  ## Downregulated genes
  temp <- apply(cnts %>% dplyr::select(all_of(downGenes)), 2, scale)
  temp$downZsum <- rowSums(temp)
  ## Add Z score to dataframe 
  cnts$downZsum <- temp$downZsum
  
  
  # Select columns
  cnts %<>% dplyr::select(metacols, ss.TotalScore:ss.DownDispersion, nondirectional_Zsum, upZsum, downZsum)
  cnts %<>% mutate(directional_Zsum = upZsum + (downZsum * -1)) %>%
    dplyr::select(-Row.names)
  
  # Return
  return(cnts)
  
  
}

# Test
temp <- calculate_signature_scores(nbl_df = nbl, 
                                   upGenes = colnames(nbl)[555:577],
                                   downGenes = colnames(nbl)[655:667], 
                                   metacols = 1:28)


```


```{r}

```





